\replaced{Fast Fourier transformation of electromagnetic data for computationally expensive kernels}{Fast time-domain electromagnetic modelling in the frequency domain}

### 2020-12-01 ManuscriptTime.tex
Revision-I (#2)
* Revision I
* Update for matplotlib 3.3.0
* Change from Geophysics to GJI template

\documentclass[extra, camera,%
    % onecolumn,   % comment for two columns
    referee,     % uncomment for double lines
    % final,       % uncomment for changes
]{gji}
\usepackage{timet}

@@

\begin{summary}
%
\deleted{Modelling three-dimensional time-domain controlled-source
electromagnetic data with frequen\-cy-domain solvers is an alternative to
time-domain solvers.} \added{Three-dimensional controlled-source
electromagnetic data is often computed directly in the domain of interest,
either in the frequency domain or in the time domain. Computing it in one
domain and transforming it via a Fourier transform to the other domain is a
viable alternative.} It requires the \replaced{evaluation}{computation} of many
\replaced{responses in the computational domain}{frequencies} if standard
Fourier transforms are used\replaced{. This}{, which} can make it prohibitively
expensive \replaced{if the kernel is time-consuming as is the case in
three-dimensional electromagnetic modelling}{in comparison with a time-domain
computation}. The speed of \replaced{such}{time-domain} modelling
\replaced{that is obtained through a transform}{with frequency-domain
computation} is defined by three key points: solver, method and implementation
of the Fourier transformation, and gridding.
%
The faster the \deleted{frequency-domain} solver, the faster
\deleted{time-domain} modelling will be. It is important that the solver is
robust over a wide range of \added{values (}frequencies\added{ or times)}.
\deleted{A three-dimensional, iterative, frequency-domain multigrid solver
satisfies this re\-quirement, but any other robust solver will do.}
%
The method should require as few \replaced{kernel evaluations}{frequencies} as
possible while remaining robust. As the frequency \added{and time}
range\added{s} span\deleted{s} many orders of magnitude, the required
\replaced{values}{frequencies} are ideally equally spaced on a logarithmic
scale. The \added{proposed} fast \deleted{frequency-to-time domain} method uses
either the digital linear filter method or the logarithmic fast Fourier
transform together with a careful selection of \replaced{evaluation
points}{frequencies} and interpolation. \replaced{In frequency-to-time domain
tests t}{T}his methodology requires typically 15 to 20 frequencies to cover a
wide range of offsets\deleted{in one-dimensional test models}.
%
The gridding should be frequency-\added{ or time-}dependent, which is
accomplished by making it a function of skin depth. Optimising for the least
number of required cells \deleted{for a given frequency} should be combined
with optimising for computational speed.
%
Looking carefully at these \deleted{three} points results in much smaller
computation times with speedup factors of ten or more over previous results. A
\deleted{frequency-domain} computation \replaced{followed by transformation}{of
time-domain data} can therefore be competitive to \replaced{computation in the
required }{time-}domain \deleted{solvers} if the required \replaced{evaluation
points}{frequencies} and the corresponding grids are carefully chosen.
%
\end{summary}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or geological purposes \citep{NAT.19.Johanson}. While
current sources with a few frequencies are used in the deep marine environment,
transient measurements are more common in the shallow marine environment and on
land \citep[e.g., ][]{GEO.07.Ziolkowski, SEG.07.Andreis, SEG.07.Avdeeva}. One
of the main reasons is the dominance of the airwave in shallow marine and
terrestrial measurements, which can be better separated in the time domain.
CSEM is usually divided into frequency- and time-domain methods, depending
whether the source signal is a continuous waveform, such as a sine, or a finite
waveform, such as a pseudo-random binary sequence (PRBS). A numerical
comparison of the two methods is given by \cite{GP.13.Conell}. Acquired CSEM
data are subsequently often analysed (modelled and inverted for) in their
respective domain, either the frequency or time domain. Modellers of layered
media usually exploit the horizontal shift-invariance by computing the
responses in the wavenumber-frequency domain followed by a 2D inverse spatial
Fourier transform, also called Hankel transform, to the space-frequency domain,
and a regular inverse Fourier transform if time-domain data are required
\citep[e.g., ][]{GEO.15.Hunziker}.

CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time. There is a
wealth of 3D electromagnetic codes, with a first boom in the 1990's, with a
good state of the art provided by \cite{B.SEG.99.Oristaglio}. A second wave
occurred during the CSEM boom in the hydrocarbon industry in the early 21th
century, see, e.g., \cite{SG.05.Avdeev} and \cite{SG.10.Borner} for
comprehensive overviews. \replaced{Focusing on r}{R}ecent publications
\added{after these review articles} span\added{s} the width of numerical
methods from integral equation codes \citep{MGS.17.Kruglyakov} and differential
equation methods such as finite differences
\replaced{\mbox{\citep{GEO.10.Mittet, GEO.11.Zaslavsky, CAG.13.Sommer}}}
{\mbox{\citep{CAG.13.Sommer}}}, finite elements \citep{GJI.13.Grayver}, and
finite volumes \citep{GEO.14.Jahandari}. Besides the different methods there
are also many different types of discretisation. Commonly used, besides
rectilinear Yee grids \citep{IEEE.66.Yee} as used by our code, are Lebedev
grids \citep{CMMP.64.Lebedev} but also unstructured grids using, for instance,
tetrahedral \citep{CAG.17.Cai}. The computational requirement to model CSEM
data also depends heavily on the technique which is used to solve the system,
where there are direct solvers \citep{GEO.15.Grayver}, indirect solvers
\citep{GJI.15.Jaysaval}, and more recently also hybrid solvers
\citep{GEO.18.Liu}.

Under certain conditions it can be \replaced{a viable alternative}{competitive}
to model time-domain data with a frequency-domain code, as shown by
\cite{GEO.08.Mulder}. \added{They compared the computational complexity of
explicit time-stepping schemes (Du Fort-Frankel method as used in, e.g.,
\mbox{\cite{GEO.04.Commer, GEO.07.Maao}}); implicit schemes \mbox{\citep[e.g.,
][]{GEO.04.Haber}}, and matrix exponentials and Lanczos reduction schemes
\mbox{\cite[e.g., ][]{RS.94.Druskin}} with the Fourier transform of
frequency-domain solutions. Their conclusion was that the Fourier transform
method can be favourable if the number of required frequencies is small
relative to the number of grid points in each coordinate.} The conditions
\added{for fast computation of time-domain data with a frequency-domain code}
are: a sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are an adaptive, frequency-dependent gridding
scheme that minimizes the required cells in each dimension, and a logarithmic
Fourier transform such as digital linear filters \citep[DLF, ][]{GP.71.Ghosh}
or the logarithmic fast Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton}
to go from the frequency to the time domain. The latter makes it also possible
to only use the imaginary part of the frequency-domain response, which has
advantages when it comes to interpolation.

\added{While we use the finite-integration technique
\mbox{\citep{AEU.77.Weiland}} in the frequency domain to show our method it is
important to note that our ideas can be used with any solver, it is not bound
to one or another frequency- or time-domain solver. The ideas for the frequency
selection can be applied to time selection, and time-dependent, adaptive
gridding exists as well \mbox{\citep[e.g., ][]{RS.06.Commer}}.} In the next
section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of our changes to the Fourier
transform and the adaptive gridding. Finally, we demonstrate the efficiency of
the approach with some numerical results.

@@

Being able to model CSEM data both in the frequency-domain and in the
time-domain can be desirable, as both domains have advantages and
disadvantages. One way to achieve this is to implement Maxwell's equation in
both domains\deleted{, as it is done, for instance, in \simpeg (Cockett et al.,
2015)}. Another approach is to have Maxwell's equation only implemented in one
domain, and use Fourier transforms to go to the other. However, this approach
can be costly, as many frequencies over a wide range are required to go from
the time domain to the frequency domain, or many times over a wide range for
the opposite direction. We present a methodology which significantly reduces
the amount and range of the required frequencies, translating into a
significant reduction in computation time. The required computation grids for
low frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain is naturally based on the skin depth,
which is the distance after which the amplitude of the electromagnetic field is
decayed by $1/e\approx 37\,\%$. The skin depth $\delta$ is a function of
conductivity and frequency, and for the diffusive approximation of Maxwell's
equation in an isotropic, homogeneous medium is given by \citep[e.g.,][
equation~1.53]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \
  % \frac{503.3}{\sqrt{f\sigma}} \, ,
  % 503.3(f\sigma)^{-1/2} \, ,
  503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

\cite{GEO.07.Plessix} define the minimum cell width $\Delta_\mathrm{min}$ as a
fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be 2--3 times bigger than $\Delta_\mathrm{min}$.
The cells have to be smallest around the source; in the marine case, the
minimum skin depth is therefore defined by the conductivity of seawater.
However, this yields quite large cells for low frequencies, so special care has
to be taken around the source by defining a maximum allowed
$\Delta_\mathrm{min}$. The grid dimension, on the other hand, is defined as a
function of the skin depth for the average conductivity of the background,
$\delta_\mathrm{ave}$. They use four times $\delta_\mathrm{ave}$ for the $x$-,
$y$-, and downward $z$-directions, and a fixed 50\,km for the upward
$z$-direction to account for the airwave. To reduce the number of cells, it is
desirable to introduce stretching, at least in the boundary zone outside of the
area where source and receivers are located.

\cite{GEO.08.Mulder} provide a computational complexity analysis of various
methods to model transient electromagnetic responses directly in the
time-domain, and compare it to the computation of transient EM responses in the
frequency-domain with a subsequent Fourier transform. They conclude the review
by stating \emph{\guillemotleft Although it remains to be seen which of the
four methods requires the least computer time for a given accuracy, the
frequency-domain approach appears to be attractive.\guillemotright} Their
approach is to minimize the computation time by having, in addition to the just
introduced adaptive gridding, an adaptive frequency selection scheme. This
scheme starts with computing the responses for a set of just five frequencies,
regularly sampled on a log-scale, from minimum to maximum required frequency.
All the other frequencies are interpolated with a shape-preserving
piecewise-cubic Hermite interpolation \citep[PCHIP, ][]{SIAM.80.Fritsch}.
Testing the stability of the obtained response by removing a single
frequency-value at a time their scheme decides if more frequencies in-between
the already computed ones are required. In this way frequencies are only added
if required, hence if certain criteria of response stability are not met. While
this method is good and effective for a single offset, it loses all its
advantages if one tries to compute different offsets within one computation, as
each offset requires a different set of adaptive frequencies. Additionally, it
hampers the parallelization over frequencies.

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code even more competitive in comparison with time-domain
codes. It is important to note that while we show the transform from frequency
to time domain, the conclusions regarding the Fourier transform can equally
well be applied to the transform from time to frequency domain data for a code
that solves Maxwell's equations in the time domain. A variation of our method
was implemented, for instance, by \custem \citep{GEO.19.Rochlitz} to obtain
frequency-domain data from time-domain computations (Rochlitz, private
communication, 2020).

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \deleted{\emg3d has implemented }known improvements such as
line-relaxation and semicoarsening \citep{ECCFD.06.Jonsthovel} \added{are
implemented}, with a non-standard Cholesky decomposition to speed up the
computation \citep{GEO.08.Mulder}. One of the big advantages of the multigrid
method is that it scales linearly (optimal) with the grid size in both CPU and
RAM usage \citep{B.Springer.20.Mulder}. This makes it feasible to run even big
models on standard computers, without the need for big clusters. All the
examples in this article are run on a laptop with an i7-6600U\,CPU\,@\,2.6\,GHz
(x4) and 16\,GB of memory, using Ubuntu \replaced{20.04}{18.04} and Python
\replaced{3.8}{3.7}. \added{All examples were run using a single thread, but
parallelization over frequencies is straight forward, as the computations are
independent.} For the semi-analytical computations of layered models we use the
open-source (Apache License 2.0) code \empymod \citep{GEO.17.Werthmuller}.

\section{Frequency selection} % % % % % % % % % % % % % % % % % % % % % % % % %

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. \added{The same applies for the reverse operation, the
fewer times required, the quicker we obtain frequency-domain results. Even
though we limit our analysis here to frequency-to-time transformation, most of
the arguments are reversible and applicable to time-to-frequency
transformations.} We decided to use a regular spacing of frequencies on a
log-scale, rather than an adaptive scheme. This approach is favourable if a
wide range of offsets is needed, as the required frequencies change with offset
and an adaptive frequency selection is therefore often tailored to a single
offset. Also, it allows for straightforward parallelization over frequencies,
which is not completely possible with an adaptive scheme.

@@

Figure~\ref{fig:marine} shows the result for an offset of 5\,km, in (a) the
frequency and (b) the time domain. The recovered response with the 3D code
captures the airwave (first peak) and the subsurface (second peak) very
accurately. At later times the error starts to increase. We would need to
compute a few additional lower frequencies if we want to improve it.
%
\begin{figure*}
  \centering
  \includegraphics[width=\fwidth]{06-marine}
  \caption{Electric inline response at an offset of 5\,km for a shallow marine,
    layered scenario. (a) Frequency-domain, where the blue circles are computed
    responses and the black dots are interpolated responses or responses set to
    zero. (b) Time-domain response.}
  \label{fig:marine}
\end{figure*}
%
In the frequency-domain plot, it can be seen that the high frequencies are not
computed very accurately, but without too much influence on the time-domain
response. These frequencies could be left out if an offset of 5\,km is the only
objective. However, we also want to retrieve shorter offsets from the same
computation, for which these frequencies are required.

Figure~\ref{fig:marine-multioffset} shows the time-domain responses of the same
model for offsets of 3, 5, and 7\,km, all obtained with the same
frequency-domain computations and the same frequencies for the Fourier
transform. The computation of these frequencies took less than 13 minutes and
it handles any offset between 3 and 7\,km. It can be seen that the chosen
frequency selection is sufficient for this offset range; again, more low
frequencies could be added to improve late-time values.
%
\begin{figure}
  \centering
  \includegraphics[width=\cwidth]{07-marine-multioffset}
  \caption{Time-domain responses for offsets of 3, 5, and 7\,km for the same
    model as shown in Figure~\ref{fig:marine}.}
  \label{fig:marine-multioffset}
\end{figure}
%

@@

The skin-depth approach fails for the air layer, as explained in the Section
\emph{Gridding}. The reason is that the EM field in the air travels at the
speed of light as a wave, and its amplitude is only reduced through geometrical
spreading. On land and in shallow marine scenarios one has therefore to include
a sufficiently large computational domain, and the default in our scheme is
100\,km. The important point is that this does not only apply to the upward
$z$-direction, but also to the horizontal directions, as the airwave also
bounces back horizontally and would continuously emit energy into the
subsurface if the boundaries are not chosen far enough away from the receivers.
If models are computed with very resistive layers or models with highly
resistive basements, this can even apply to deep marine scenarios.

>>>>

\documentclass[extra, camera,%
    % onecolumn,   % comment for two columns
    % referee,     % uncomment for double lines
    final,       % uncomment for changes
]{gji}
\usepackage{timet}

@@

\begin{summary}
%
\deleted{Modelling three-dimensional time-domain controlled-source
electromagnetic data with frequen\-cy-domain solvers is an alternative to
time-domain solvers.} \added{Three-dimensional controlled-source
electromagnetic data is often computed directly in the domain of interest,
either in the frequency domain or in the time domain. Computing it in one
domain and transforming it via a Fourier transform to the other domain is a
viable alternative.} It requires the \replaced{evaluation}{computation} of many
\replaced{responses in the computational domain}{frequencies} if standard
Fourier transforms are used\replaced{. This}{, which} can make it prohibitively
expensive \replaced{if the kernel is time-consuming as is the case in
three-dimensional electromagnetic modelling}{in comparison with a time-domain
computation}. The speed of \replaced{such}{time-domain} modelling
\replaced{that is obtained through a transform}{with frequency-domain
computation} is defined by three key points: solver, method and implementation
of the Fourier transformation, and gridding.
%
The faster the \deleted{frequency-domain} solver, the faster
\deleted{time-domain} modelling will be. It is important that the solver is
robust over a wide range of \added{values (}frequencies\added{ or times)}.
\deleted{A three-dimensional, iterative, frequency-domain multigrid solver
satisfies this re\-quirement, but any other robust solver will do.}
%
The method should require as few \replaced{kernel evaluations}{frequencies} as
possible while remaining robust. As the frequency \added{and time}
range\added{s} span\deleted{s} many orders of magnitude, the required
\replaced{values}{frequencies} are ideally equally spaced on a logarithmic
scale. The \added{proposed} fast \deleted{frequency-to-time domain} method uses
either the digital linear filter method or the logarithmic fast Fourier
transform together with a careful selection of \replaced{evaluation
points}{frequencies} and interpolation. \replaced{In frequency-to-time domain
tests t}{T}his methodology requires typically 15 to 20 frequencies to cover a
wide range of offsets\deleted{in one-dimensional test models}.
%
The gridding should be frequency-\added{ or time-}dependent, which is
accomplished by making it a function of skin depth. Optimising for the least
number of required cells \deleted{for a given frequency} should be combined
with optimising for computational speed.
%
Looking carefully at these \deleted{three} points results in much smaller
computation times with speedup factors of ten or more over previous
\replaced{methods}{results}. A \deleted{frequency-domain} computation
\replaced{in one domain followed by transformation}{of time-domain data} can
therefore be \replaced{an alternative}{competitive} to \replaced{computation in
the other domain }{time-}domain \deleted{solvers} if the required
\replaced{evaluation points}{frequencies} and the corresponding grids are
carefully chosen.
%
\end{summary}

@@

The controlled-source electromagnetic (CSEM) method is one of the common
non-seismic tools in exploration geophysics, not only in hydrocarbon
exploration \citep{GEO.10.Constable}, but also in the search for sulfides
\citep{GRL.19.Gehrmann}, water \citep{GEO.05.Pedersen}, geothermal sources
\citep{WGC.15.Girard}, or \added{for} geological purposes
\citep{NAT.19.Johanson}. While current sources with a few frequencies are used
in the deep marine environment, transient measurements are more common in the
shallow marine environment and on land \citep[e.g., ][]{GEO.07.Ziolkowski,
SEG.07.Andreis, SEG.07.Avdeeva}. One of the main reasons is the dominance of
the airwave in shallow marine and terrestrial measurements, which can be better
separated in the time domain. CSEM is usually divided into frequency- and
time-domain methods, depending \added{on} whether the source signal is a
continuous waveform, such as a sine, or a finite waveform, such as a
pseudo-random binary sequence (PRBS). A numerical comparison of the two methods
is given by \cite{GP.13.Conell}. Acquired CSEM data are subsequently often
analysed (modelled and inverted for) in their respective domain, either the
frequency or time domain. Modellers of layered media usually exploit the
horizontal shift-invariance by computing the responses in the
wavenumber-frequency domain followed by a 2D inverse spatial Fourier transform,
also called Hankel transform, to the space-frequency domain, and a regular
inverse Fourier transform if time-domain data are required \citep[e.g.,
][]{GEO.15.Hunziker}.

% a good review of the state of art 
CSEM codes for 2D and 3D computations, on the other hand, mostly compute their
responses directly in the required domain, either frequency or time. There is a
wealth of 3D electromagnetic codes, with a first boom in the 1990's\replaced{;
  \mbox{\cite{B.SEG.99.Oristaglio}} provide a good review of the state of
art at that time}{, with a good state of the art provided by
\mbox{\cite{B.SEG.99.Oristaglio}}}. A second wave occurred during the CSEM boom
in the hydrocarbon industry in the early 21th century, see, e.g.,
\cite{SG.05.Avdeev} and \cite{SG.10.Borner} for comprehensive overviews. Recent
publications \added{after these review articles} span the width of numerical
methods from integral equation codes \citep{MGS.17.Kruglyakov} and differential
equation methods such as finite differences
\replaced{\mbox{\citep{GEO.10.Mittet, GEO.11.Zaslavsky, CAG.13.Sommer}}}
{\mbox{\citep{CAG.13.Sommer}}}, finite elements \citep{GJI.13.Grayver}, and
finite volumes \citep{GEO.14.Jahandari}. Besides the different methods there
are also many different types of discretisation. Commonly used, besides
rectilinear Yee grids \citep{IEEE.66.Yee} as used by our code, are Lebedev
grids \citep{CMMP.64.Lebedev} but also unstructured grids using, for instance,
tetrahedra\deleted{l} \citep{CAG.17.Cai}. The computational requirement to
model CSEM data also depends heavily on the technique \deleted{which is} used
to solve the system\replaced{. T}{, where t}here are direct solvers
\citep{GEO.15.Grayver}, indirect solvers \citep{GJI.15.Jaysaval}, and more
recently also hybrid solvers \citep{GEO.18.Liu}.

Under certain conditions it can be \replaced{a viable alternative}{competitive}
to model time-domain data with a frequency-domain code, as shown by
\cite{GEO.08.Mulder}. \added{They compared the computational complexity of
explicit time-stepping schemes (Du Fort-Frankel method as used in, e.g.,}
\cite{GEO.04.Commer, GEO.07.Maao}\added{); implicit schemes \mbox{\citep[e.g.,
][]{GEO.04.Haber}}, and matrix exponentials and Lanczos reduction schemes}
\cite[e.g., ][]{RS.94.Druskin} \added{with the Fourier transform of
frequency-domain solutions. Their conclusion was that the Fourier transform
method can be favourable if the number of required frequencies is small
relative to the number of grid points in each coordinate.} The conditions
\added{for fast computation of time-domain data with a frequency-domain code}
are: a sufficiently powerful solver, appropriate frequency selection and
interpolation, and an automated gridding, for which they used the
multi-frequency CSEM approach presented by \cite{GEO.07.Plessix}. We build upon
these results but improve the run time from hours to minutes. The main reasons
for this significant speed-up are an adaptive, frequency-dependent gridding
scheme that minimizes the required cells in each dimension, and a logarithmic
Fourier transform such as digital linear filters \citep[DLF, ][]{GP.71.Ghosh}
or the logarithmic fast Fourier transform \citep[FFTLog, ][]{RAS.00.Hamilton}
to go from the frequency to the time domain. The latter makes it also possible
to only use the imaginary part of the frequency-domain response, which has
advantages when it comes to interpolation.

\added{While we use the finite-integration technique
\mbox{\citep{AEU.77.Weiland}} in the frequency domain to demonstrate our
method, it is important to note that our ideas can be used with any solver and
is not bound to one or another frequency- or time-domain solver. The ideas for
the frequency selection can be applied to time selection, and time-dependent,
adaptive gridding exists as well \mbox{\citep[e.g., ][]{RS.06.Commer}}.} In the
next section, we briefly review the methodology as introduced by
\cite{GEO.07.Plessix} and \cite{GEO.08.Mulder} and highlight their advantages
and shortcomings. This is followed by an outline of our changes to the Fourier
transform and the adaptive gridding. Finally, we demonstrate the efficiency of
the approach with some numerical results.

@@

Being able to model CSEM data both in the frequency\replaced{ }{-}domain and in
the time\replaced{ }{-}domain can be desirable, as both domains have advantages
and disadvantages. One way to achieve this is to implement Maxwell's
equation\added{s} in both domains\deleted{, as it is done, for instance, in
\simpeg (Cockett et al., 2015)}. Another approach is to have Maxwell's
equation\added{s} only implemented in one domain, and use Fourier transforms to
go to the other. However, this approach can be costly, as many frequencies over
a wide range are required to go from the \replaced{frequency}{time} domain to
the \replaced{time}{frequency} domain, or many times over a wide range for the
opposite direction. We present a methodology which significantly reduces the
amount and range of the required frequencies, translating into a significant
reduction in computation time. The required computation grids for low
frequencies (in our case in the order of 0.001\,Hz) and high frequencies
(around 100\,Hz) are hugely different. Low frequencies can be computed on a
coarser grid, but they require a much larger domain in order to avoid boundary
effects. High frequencies, on the other hand, require denser gridding, but they
are much more limited in reach. An adaptive gridding scheme is therefore
required. We build our approach upon \cite{GEO.07.Plessix}, who presented such
an adaptive gridding for multi-frequency (and multi-source) CSEM modelling.

An adaptive grid in the frequency domain is naturally based on the skin depth,
which is the distance after which the amplitude of the electromagnetic field is
decayed by $1/e\approx 37\,\%$. The skin depth $\delta$ is a function of
conductivity and frequency, and for the diffusive approximation of Maxwell's
equation\added{s} in an isotropic, homogeneous medium is given by
\citep[e.g.,][ equation~1.53]{B.SEG.88.Ward}
%
\begin{equation}
  \delta = \sqrt{\frac{2}{\omega\mu\sigma}}
  \ \approx \
  % \frac{503.3}{\sqrt{f\sigma}} \, ,
  % 503.3(f\sigma)^{-1/2} \, ,
  503.3/\sqrt{f\sigma} \, ,
  \label{eq:skindepth}
\end{equation}
%
where $\sigma$ is conductivity (S/m), $\omega=2\pi f$ is angular frequency of
frequency $f$ (Hz), and $\mu$ is magnetic permeability (H/m). The approximation
is obtained by using the free-space value of magnetic permeability,
$\mu_0=4\pi\times10^{-7}\,$H/m.

\cite{GEO.07.Plessix} define the minimum cell width $\Delta_\mathrm{min}$ as a
fraction of the minimum skin depth $\delta_\mathrm{min}$, where
$\delta_\mathrm{min}$ should be 2--3 times bigger than $\Delta_\mathrm{min}$.
The cells have to be smallest around the source; in the marine case, the
minimum skin depth is therefore defined by the conductivity of seawater.
However, this yields quite large cells for low frequencies, so special care has
to be taken around the source by defining a maximum allowed
$\Delta_\mathrm{min}$. The grid dimension, on the other hand, is defined as a
function of the skin depth for the average conductivity of the background,
$\delta_\mathrm{ave}$. They use four times $\delta_\mathrm{ave}$ for the $x$-,
$y$-, and downward $z$-directions, and a fixed 50\,km for the upward
$z$-direction to account for the airwave. To reduce the number of cells, it is
desirable to introduce stretching, at least in the \replaced{buffer}{boundary}
zone outside of the area where source and receivers are located.

\cite{GEO.08.Mulder} provide a computational complexity analysis of various
methods to model transient electromagnetic responses directly in the
time\replaced{ }{-}domain, and compare it to the computation of transient EM
responses in the frequency\replaced{ }{-}domain with a subsequent Fourier
transform. They conclude the review by stating \emph{\guillemotleft Although it
remains to be seen which of the four methods requires the least computer time
for a given accuracy, the frequency-domain approach appears to be
attractive.\guillemotright} Their approach is to minimize the computation time
by having, in addition to the just introduced adaptive gridding, an adaptive
frequency selection scheme. This scheme starts with computing the responses for
a set of just five frequencies, regularly sampled on a log-scale, from minimum
to maximum required frequency. All the other frequencies are interpolated with
a shape-preserving piecewise-cubic Hermite interpolation \citep[PCHIP,
][]{SIAM.80.Fritsch}. Testing the stability of the obtained response by
removing a single frequency-value at a time their scheme decides if more
frequencies in-between the already computed ones are required. In this way
frequencies are only added if required, hence if certain criteria of response
stability are not met. While this method is good and effective for a single
offset, it loses all its advantages if one tries to compute different offsets
within one computation, as each offset requires a different set of adaptive
frequencies. Additionally, it hampers the parallelization over frequencies.

We present improvements to both the adaptive gridding and the transform from
frequency domain to time domain, which makes time-domain modelling with a
frequency-domain code \replaced{viable}{even more competitive} in comparison
with time-domain codes. It is important to note that while we show the
transform from frequency to time domain, the conclusions regarding the Fourier
transform can equally well be applied to the transform from time\added{-domain}
to frequency\replaced{-}{ }domain data for a code that solves Maxwell's
equations in the time domain. A variation of our method was implemented, for
instance, by \custem \citep{GEO.19.Rochlitz} to obtain frequency-domain data
from time-domain computations (Rochlitz, private communication, 2020).

@@

For the numerical computations we use the open-source (Apache License 2.0) code
\emg3d \citep{JOSS.19.Werthmuller}, a multigrid solver based on
\cite{GP.06.Mulder}, which can be used as a preconditioner for Krylov subspace
solvers or as a solver on its own. The multigrid approach works fine for the
diffusive approximation of Maxwell's equation\added{s}, which assumes that
$\omega\varepsilon \ll \sigma$, where $\varepsilon$ is electric permittivity
(F/m). The remaining system to solve in the frequency domain is then given by
the second-order differential equation of the electric field,
%
\begin{equation}
    \mathrm{i}\omega\sigma \mathbf{E} +
    \nabla \times \mu^{-1} \nabla \times \mathbf{E}
    = -\mathrm{i}\omega\mathbf{J}_\mathrm{s} \, ,
  \label{eq:maxwell}
\end{equation}
%
where $\mathbf{E}$ is the electric field (V/m) and $\mathbf{J}_\mathrm{s}$ the
current source (A/m$^2$); time dependence is $\exp(\mathrm{i}\omega t)$. The
standard multigrid approach fails for severe stretching or strong anisotropy,
for which \deleted{\emg3d has implemented }known improvements such as
line-relaxation and semicoarsening \citep{ECCFD.06.Jonsthovel} \added{are
implemented}, with a non-standard Cholesky decomposition to speed up the
computation \citep{GEO.08.Mulder}. One of the big advantages of the multigrid
method is that it scales linearly (optimal) with the grid size in both CPU and
RAM usage \citep{B.Springer.20.Mulder}. This makes it feasible to run even big
models on standard computers, without the need for big clusters. All the
examples in this article are run on a laptop with an i7-6600U\,CPU\,@\,2.6\,GHz
(x4) and 16\,GB of memory, using Ubuntu \replaced{20.04}{18.04} and Python
\replaced{3.8}{3.7}. \added{All examples were run using a single thread, but
parallelization over frequencies is straight forward, as the computations are
independent.} For the semi-analytical computations of layered models we use the
open-source (Apache License 2.0) code \empymod \citep{GEO.17.Werthmuller}.

\subsection{Frequency selection}  % % % % % % % % % % % % % % % % % % % % % % %

An important factor in terms of speed and accuracy for time-domain responses
obtained from frequency-domain computations is the selection of the required
frequencies. The fewer frequencies required, the quicker we obtain the
time-domain result. \added{The same applies for the reverse operation, the
fewer times required, the quicker we obtain frequency-domain results. Even
though we limit our analysis here to frequency-to-time transformation\added{s},
most of the arguments are reversible and applicable to time-to-frequency
transformations.} We decided to use a regular spacing of frequencies on a
log-scale, rather than an adaptive scheme. This approach is favourable if a
wide range of offsets is needed, as the required frequencies change with offset
and an adaptive frequency selection is therefore often tailored to a single
offset. Also, it allows for straightforward parallelization over frequencies,
which is not completely possible with an adaptive scheme.

@@

Figure~\ref{fig:marine} shows the result for an offset of 5\,km, in (a) the
frequency and (b) the time domain. The recovered response with the 3D code
captures the airwave (first peak) and the subsurface (second peak) very
accurately. At later times the error starts to increase. We would need to
compute a few additional lower frequencies if we want to improve it.
%
\begin{figure*}
  \centering
  \includegraphics[width=\fwidth]{06-marine}
  \caption{Electric inline response at an offset of 5\,km for a shallow marine,
  layered scenario. (a) Frequency-domain\added{ response}, where the blue
  circles \replaced{denote}{are} computed responses and the black dots
  \deleted{are} interpolated responses or responses set to zero. (b)
  Time-domain response.}
  \label{fig:marine}
\end{figure*}
%
In the frequency-domain plot, it can be seen that the high frequencies are not
computed very accurately, but without too much influence on the time-domain
response. These frequencies could be left out if an offset of 5\,km is the only
objective. However, we also want to retrieve shorter offsets from the same
computation, for which these frequencies are required.

Figure~\ref{fig:marine-multioffset} shows the time-domain responses of the same
model for offsets of 3, 5, and 7\,km, all obtained with the same
frequency-domain computations and the same frequencies for the Fourier
transform. The computation of these frequencies took less than \replaced{9}{13}
minutes and it handles any offset between 3 and 7\,km. It can be seen that the
chosen frequency selection is sufficient for this offset range; again, more low
frequencies could be added to improve late-time values.
%
\begin{figure}
  \centering
  \includegraphics[width=\cwidth]{07-marine-multioffset}
  \caption{Time-domain responses for offsets of 3, 5, and 7\,km for the same
    model as shown in Figure~\ref{fig:marine}.}
  \label{fig:marine-multioffset}
\end{figure}
%

@@

The skin-depth approach fails for the air layer, as explained in the Section
\emph{Gridding}. The reason is that the EM field in the air travels at the
speed of light as a wave, and its amplitude is only reduced through geometrical
spreading. On land and in shallow marine scenarios one has therefore to include
a sufficiently large computational domain\replaced{. T}{, and t}he default in
our scheme is 100\,km. The important point is that this does not only apply to
the upward $z$-direction, but also to the horizontal directions, as the airwave
\deleted{also} bounces back horizontally and would continuously emit energy
into the subsurface if the boundaries are not chosen far enough away from the
receivers. If models are computed with very resistive layers or models with
highly resistive basements, this can even apply to deep marine scenarios.
